---
title: "Fuel Efficiency Analysis of Cars Using R and Python"
author: "Gabriella Angelica Jemima Hasian"
date: "2026-09-24"
---

## Introduction

This analysis examines fuel efficiency in the `mtcars` dataset while demonstrating how R and Python can work together within the same Quarto document. R is used to prepare the data and calculate average fuel efficiency by cylinder group. The resulting R objects are then passed to Python for further analysis and visualization. Finally, a result calculated in Python is passed back to R.

The `mtcars` dataset contains fuel consumption and other characteristics of 32 automobiles. The dataset is included in R's `datasets` package and was originally extracted from the 1974 issue of *Motor Trend* magazine.

**Data source:** R `datasets` package — `mtcars` documentation:  
<https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html>

## Setting Up R and Python

I use `reticulate` to connect R and Python within the same Quarto document. Python is configured to use the project's `.venv` environment so that the same environment used by the rest of the project is used during rendering.

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(reticulate)

use_virtualenv("../../.venv", required = TRUE)

py_config()
```

## Preparing the Data in R

I begin with the built-in `mtcars` dataset in R. I move the car names from the row names into a regular column and convert the number of cylinders into a categorical variable. This makes the cylinder groups easier to compare in the analysis.

```{r}
mtcars_data <- mtcars |>
  rownames_to_column("model") |>
  mutate(cyl = factor(cyl))

head(mtcars_data)
```

## Calculating Fuel Efficiency by Cylinder Group in R

I calculate the average miles per gallon (`mpg`) and number of cars for each cylinder group.

The resulting `r_summary` object is calculated entirely in R and will be passed to Python in the next section.

```{r}
r_summary <- mtcars_data |>
  group_by(cyl) |>
  summarise(
    mean_mpg = mean(mpg),
    n = n(),
    .groups = "drop"
  ) |>
  mutate(
    mean_mpg = round(mean_mpg, 2)
  )

r_summary
```

The results show how average fuel efficiency differs across the cylinder groups. I now pass this R object to Python for further analysis.

## Using R Objects in Python

The `r` object provided by `reticulate` allows Python to access objects created in R. Here, Python receives both the car-level dataset and the summary calculated in R.

```{python}
#| warning: false
#| message: false

import pandas as pd
import matplotlib.pyplot as plt

python_data = r.mtcars_data
python_summary = r.r_summary

python_summary
```

The `python_summary` object is therefore not independently created in Python. It is the `r_summary` object that was calculated in R and passed into the Python session.

## Identifying the Highest-Fuel-Efficiency Group in Python

I use Python to identify the cylinder group with the highest average fuel efficiency.

```{python}
top_cyl = python_summary.loc[
    python_summary["mean_mpg"].idxmax(),
    "cyl"
]

top_mpg = python_summary["mean_mpg"].max()

print("Cylinder group with the highest mean MPG:", top_cyl)
print("Highest mean MPG:", top_mpg)
```

The Python calculation identifies the cylinder group with the highest average MPG. The resulting `top_cyl` object will now be passed back to R.

## Passing the Python Result Back to R

The `py` object provided by `reticulate` allows R to access objects created in Python. I use `py$top_cyl` to retrieve the result calculated by Python.

```{r}
top_cyl_from_python <- py$top_cyl

top_cyl_from_python
```

This demonstrates the two-way exchange of objects between the two languages:

```text
R object → Python → Python result → R
```

## Visualizing Fuel Efficiency in Python

The following visualization uses the car-level data originally prepared in R and passed to Python as `python_data`.

The x-axis represents vehicle weight, while the y-axis represents fuel efficiency. Point size represents horsepower, and the legend distinguishes cars by cylinder group. Individual car names are also labelled to make the observations identifiable.

```{python}
#| label: fig-mtcars-analysis
#| fig-cap: "Vehicle weight and fuel efficiency across cars, with point size representing horsepower and cylinder group represented by the legend"

fig, ax = plt.subplots(figsize=(10, 6))

for cyl in sorted(python_data["cyl"].unique()):
    subset = python_data[python_data["cyl"] == cyl]

    _ = ax.scatter(
        subset["wt"],
        subset["mpg"],
        s=subset["hp"] * 1.8,
        alpha=0.75,
        label=f"{int(cyl)} cylinders",
    )

    for _, row in subset.iterrows():
        _ = ax.annotate(
            row["model"],
            (row["wt"], row["mpg"]),
            xytext=(5, 5),
            textcoords="offset points",
            fontsize=7,
            alpha=0.8,
        )

_ = ax.set_xlabel("Vehicle weight (1000 lbs)")
_ = ax.set_ylabel("Fuel efficiency (MPG)")
_ = ax.set_title("Vehicle Weight vs. Fuel Efficiency")
_ = ax.legend(title="Cylinder group")
_ = ax.grid(alpha=0.2)

_ = plt.tight_layout()
plt.show()
```

The visualization shows the relationship between vehicle weight and fuel efficiency while providing additional information about horsepower and cylinder group. The data used for the plot originated in R and was passed to Python through `reticulate`.

## Conclusion

This analysis demonstrates how R and Python can be combined within a single Quarto document using `reticulate`.

R was used to prepare the `mtcars` dataset and calculate average MPG by cylinder group. The resulting `r_summary` object was passed from R to Python using `r.r_summary`. Python then used this object to identify the cylinder group with the highest average MPG.

The Python result was subsequently passed back to R using `py$top_cyl`. Python was also used to visualize the car-level data that had originally been prepared in R.